Taking Dyads Seriously

### 2018-02-12 paper/2_model.tex
more cleanup

These types of patterns frequently emerge in IR contexts as well.\footnote{For example, see: \citet{manger:etal:2012, kinne:2013, chyzh:2016}.} For example, a perennial finding in the interstate trade literature emphasizes the role that geography plays in determining trade flows. Geographic proximity in the network context is an example of a shared attribute between actors that corresponds to a greater likelihood of the event of interest taking place. Alternatively, in the interstate conflict literature, we may find that actors who are each a member of a particular (formal or informal) alliance are likely to act similarly in the conflict network. Specifically, they will tend to initiate conflictual events with actors that their fellow alliance members initiate conflict with, and they will be unlikely to initiate conflict with members of their alliance. In both these examples, we are able to explicitly parameterize the attribute that might explain the emergence of higher order dependence patterns, but it is likely that at times we will not able to account for the variety of reasons why higher order dependence patterns in networks may develop.

>>>>

These types of patterns frequently emerge in IR contexts as well.\footnote{For example, see: \citet{manger:etal:2012, kinne:2013, chyzh:2016}.} For example, a perennial finding in the interstate trade literature emphasizes the role that geography plays in determining trade flows. Geographic proximity in the network context is an example of homophily --- a shared attribute between actors that corresponds to a greater likelihood of the event of interest taking place. Alternatively, in the interstate conflict literature, we may find that actors who are each a member of a particular (formal or informal) alliance are likely to act similarly in the conflict network. Specifically, they will tend to initiate conflictual events with actors that their fellow alliance members initiate conflict with, and they will be unlikely to initiate conflict with members of their alliance --- an example of stochastic equivalence. In both these examples, we are able to explicitly parameterize the attribute that might explain the emergence of higher order dependence patterns, but it is likely that at times we will not able to account for the variety of reasons why higher order dependence patterns in networks may develop.

### 2018-02-12 paper/4_app.tex
more cleanup

In order to showcase the insights the AME approach permits scholars to bring into their analysis, we apply AME to five recent studies from international relations and international political economy: \citet{reiter:stam:2003, mcdonald:2004,  rose:2004, weeks:2012, gibler:2017}. Each of these studies are representative of broader trends in the field in that they use relational data of state interactions and propose both dyadic, monadic, and structural explanations for behavior of actors in the system. We chose to demonstrate the capabilities of AME with reference to existing studies in order to highlight several features of the AME approach. First, the results of AME estimation are interpretable alongside GLM results but have the additional benefit of being able to quantify higher-order features of interest to scholars, such as ``sender'' and ``receiver'' effects. Second, through AME, we can examine the extent to which interdependencies within typical dyadic data alter previous conclusions about international relations from recent empirical studies using a framework that a priori assumes conditional independence. Third, we will show that by using the AME framework scholars can better model the data generating process behind their events of interest.

@@

First and foremost, many findings that emerge from models that do not take interdependencies into account lose their statistical significance when network effects are estimated via AME.\footnote{This should not be news---since the finding has long been in the theoretical literature, but given the state of current literature in international relations is still pertinent.} Not only are coefficients biased in the GLM approaches to the analysis of dyadic data, but they are often imprecisely measured, with poorly calibrated standard errors.  This means that significance testing (for better or worse) is compromised when network effects are ignored.

@@

Fourth, it is clear that the AME approach dominates the GLM approaches in terms of performance. Not only it is better at correctly identifying cases in which the dependent variable takes a value of $0$ (via the ROC curves and associated statistics), but it also dominates at correctly identifying occurrences of the dependent variable in the data (seen via the PR curves and associated statistics).  In the case of studies with continuous dependent variables, the AME approach has average error statistics that are about one-half that found in the OLS model. It is rare for studies in this area to provide performance statistics, but at the same time at least one of the studies is unable to identify a single case in spite of having almost a million observations.

>>>>

To showcase how the AME generates new insights for researchers, we apply AME to five recent studies from the broad field of international relations and international political economy: \citet{reiter:stam:2003, mcdonald:2004,  rose:2004, weeks:2012, gibler:2017}. Each of these studies are representative of broader trends in the field in that they use relational data of state interactions and propose both dyadic, monadic, and structural explanations for behavior of actors in the system. We chose to demonstrate the capabilities of AME with reference to existing studies in order to highlight several features of the AME approach. First, the results of AME estimation are interpretable alongside GLM results but have the additional benefit of being able to quantify higher-order features of interest to scholars, such as ``sender'' and ``receiver'' effects. Second, through AME, we can examine the extent to which interdependencies within typical dyadic data alter previous conclusions about international relations from recent empirical studies using a framework that a priori assumes conditional independence. Third, we will show that by using the AME framework scholars can better model the data generating process behind their events of interest.

@@

First and foremost, many findings that emerge from models that do not take interdependencies into account lose their statistical significance when network effects are estimated via AME. Not only are coefficients biased in the GLM approaches to the analysis of dyadic data, but they are often imprecisely measured, with poorly calibrated standard errors.  This means that significance testing (for better or worse) is compromised when network effects are ignored.

@@

Fourth, it is clear that the AME approach dominates the GLM approaches in terms of performance. Not only it is better at correctly identifying cases in which the dependent variable takes a value of $0$ (via the ROC curves and associated statistics), but it also dominates at correctly identifying occurrences of the dependent variable in the data (seen via the PR curves and associated statistics).  In the case of studies with continuous dependent variables, the AME approach has average error statistics that are about one-half that found in the OLS model.
